Cooling Requirements for the Vertical Shear Instability in Protoplanetary Disks

### 2015-04-29 linear_adia.tex
minute changes to sec 3, 4, 5, appendix A; partial re-write of appendix B; new file for fig 1 not found, please update

In \S\ref{relax_pert} we prove that even an infinitesimal increase in the cooling time,
starting from $\beta = 0$, slows the growth rate of the VSI.  In \S\ref{disp_relax} we 
derive exact solutions to the simplified VSI model developed above (Eq.\ \ref{nearly_iso_explicit}).
In \S\ref{iso_vsi_beta_crit} we reach our main result, the maximum cooling time above which 
VSI growth is suppressed.  We heuristically argue for this result in \S\ref{vsi_require} and numerically
verify it in \S\ref{numerical}.

>>>>

In \S\ref{relax_pert} we show that even an infinitesimal increase in the cooling time,
starting from $\beta = 0$, slows the growth rate of the VSI.  In \S\ref{disp_relax} we 
derive exact solutions to the simplified VSI model developed above (Eq.\ \ref{nearly_iso_explicit}).
In \S\ref{iso_vsi_beta_crit} we reach our main result, the maximum cooling time above which 
VSI growth is suppressed.  We heuristically argued for this result in
\S\ref{vsi_require}, and will numerically
verify it in \S\ref{numerical}.

### 2015-04-29 linear_iso.tex
minute changes to sec 3, 4, 5, appendix A; partial re-write of appendix B; new file for fig 1 not found, please update

Our simplified model starts with  Eqs.\ \ref{lin_all} 
and makes the following additional approximations:

@@

\item  We set $\zeta = 0$,  dropping terms with an explicit dependence on the 
radial structure of the equilibrium disk.  Vertical shear, which implicitly depends on the radial
temperature gradient, is retained.  This fully-radially-local approximation is also made in the
 `vertically global shearing box' (VGSB) of \citetalias{mcnally14}.

\item We make the low frequency approximation, $|\upsilon^2|\ll \kappa^2,\,\Omega^2$,
keeping only the lowest order terms in $|\upsilon|/\kappa$.  Similar to the 
incompressible \citepalias{goldreich67} or anelastic \citepalias{barker15} approximations,
the low frequency approximation filters p-modes in favor of g-modes, i.e.\ inertial
gravity waves \citep{lubow93}.

@@

\item We assume thin disk with $h\ll 1$ and take the $z \ll r$ limit to 
  obtain Eq.\ \ref{rhoisothin} for the equilibrium $\rho(z)$.

>>>>

Our simplified model starts with  Eqs.\ \ref{lin_all} 
and makes the following additional simplifications:

@@

\item  We set $\zeta = 0$,  neglecting terms with an explicit dependence on the 
radial structure of the equilibrium disk.  Vertical shear, which implicitly depends on the radial
temperature gradient, is retained.  This fully-radially-local approximation is also made in the
 `vertically global shearing box' of \citetalias{mcnally14}.

\item We make the low frequency approximation, assuming $|\upsilon^2|\ll \kappa^2,\,\Omega^2$. 
%keeping only the lowest order terms in $|\upsilon|/\kappa$.  
Similar to the incompressible \citepalias{goldreich67} or anelastic \citepalias{nelson13,barker15} approximations,
the low frequency approximation filters p-modes in favor of g-modes, i.e.\ inertial
gravity waves \citep{lubow93}.

@@

\item We consider thin disks with $h\ll 1$ 
  %and take the $z \ll r$ limit to 
  %obtain 
  and use the Gaussian approximation, Eq.\ \ref{rhoisothin}, for the equilibrium density field.

### 2015-04-29 paper.tex
minute changes to sec 3, 4, 5, appendix A; partial re-write of appendix B; new file for fig 1 not found, please update

On the other hand, one cannot obtain arbitrarily short cooling times
by invoking arbitrarily small scales (cf. \citealt{urpin03}, 
\citetalias{barker15}), because very small wavelengths are in the
optically thin cooling regime,  where cooling is independent of
lengthscale. Furthermore, 
%while we do not include viscosity in our
%idealized disk models, in reality 
small-scale perturbations will be 
prone to viscous damping, including the effective viscosity from turbulent motions.
While our linear, inviscid analysis does not address this issue directly, we expect that 
shorter wavelength modes will drive weaker non-linear (and non-axisymmetric) transport.

>>>>

On the other hand, one cannot obtain arbitrarily short cooling times
by invoking arbitrarily small scales (cf. \citealt{urpin03}, 
\citetalias{barker15}), because very small wavelengths are in the
optically thin cooling regime,  where cooling is independent of
lengthscale. Furthermore, 
%while we do not include viscosity in our
%idealized disk models, in reality 
small-scale perturbations will be 
prone to viscous damping, including the effective viscosity from turbulent motions.
While our linear, inviscid calculations do not address this issue directly, we expect that 
shorter wavelength modes will drive weaker non-linear (and non-axisymmetric) transport.

### 2015-04-29 setup.tex
minute changes to sec 3, 4, 5, appendix A; partial re-write of appendix B; new file for fig 1 not found, please update

We further define a modified sound speed
\begin{align}
	c_s \equiv \sqrt{\Gamma P /\rho } , 
% c_s^2\equiv % \frac{dP}{d\rho} =
%  c_{s0}^2(r)\left(\frac{\rho}{\rho_0}\right)^{\Gamma-1} =
%  \frac{\Gamma P}{\rho},  
\end{align}
which in general differs from the  isothermal, $c_\mathrm{iso} =
c_s/\sqrt{\Gamma}$, and adiabatic $c_\mathrm{ad} =
c_s\sqrt{\gamma/\Gamma}$, sound speeds.

We also introduce the characteristic scale-height, $H = c_s(r,0)/\OmK$
and disk aspect-ratio
%We take $c_{s0}(r)$ to be another power-law:  
%\begin{align}
%  c_{s0}^2(r)=c_{s00}^2\left(\frac{r}{r_0}\right)^q, 
%\end{align}
%where $c_{s00}\equiv c_{s0}(r_0)$  and $q$ is the temperature power-law index, and
%define   
\begin{align}
   h(r) \equiv %\frac{c_{s0}(r)}{v_k(r)} \equiv
  \frac{H}{r} \propto r^{(1+q)/2}
\end{align}
We are interested in thin disks with $ h \ll 1$.

\subsubsection{Density structure}
The equilibrium density field follows from vertical hydrostatic equilibrium (Eq.\ \ref{vert_equilibrium}).
The solution for $\Gamma\neq1$ is
\begin{align}\label{eqm_dens}
  &\left(\frac{\rho}{\rho_0}\right)^{\Gamma-1} = 1 +
  \frac{\left(\Gamma-1\right)}{ h^2}\left(\frac{1}{\sqrt{1+z^2/r^2}}-1\right).
\end{align}
Note that for $\Gamma > 1 +  h^2$ there is a disk surface $H_s$
where $\rho(r,H_s)=0$.  For $h \ll 1$, this surface is at $H_s/H \approx \sqrt{2/(\Gamma-1)}$.
% Note that for $\Gamma\neq1$ the disk has a surface $z=H(r)$ such that
% $\rho(r,H)=0$. If we parametrize
% \begin{align}
%   H(r)\equiv h r,
% \end{align}
% then the imposed sound-speed and disk thickness are related by 
% \begin{align}
%   h^2(r) = \left[1-\frac{ h^2(r)}{\Gamma-1}\right]^{-2}-1. 
% \end{align}

The vertically isothermal case, $\Gamma = 1$, can be calculated either 
as a special case or as the $\Gamma \rightarrow 1$ limit of Eq. \ref{eqm_dens} to be 
\begin{subequations}\begin{align}
  \frac{\rho(r,z)}{\rho_0(r)} &=
  \exp{\left[\frac{1}{ h^2}\left(\frac{1}{\sqrt{1+z^2/r^2}}-1\right)\right]},   \\
  & \approx \exp {\left[ - {z^2 \over 2 H^2}\right]} \label{rhoisothin},
  \end{align}\end{subequations}
  where Eq.\ \ref{rhoisothin} takes the $z \ll r$ thin disk limit.  We will primarily focus on
vertically isothermal disks, the relevant case for passively irradiated
PPDs \citep{chiang97}.

 \subsubsection{Rotation and vertical shear profiles}\label{vshear_def}
The equilibrium rotation field, $\Omega(r,z)$ follows from the density field and centrifugal balance (Eq.\ \ref{hori_equilibrium}),
giving 
\begin{align}\label{eqm_rot}
  \frac{\Omega^2}{\OmK^2}=1 +
  \frac{p+q}{\Gamma} h^2(r) 
  +\frac{s}{\Gamma} \left(1-\frac{1}{\sqrt{1+z^2/r^2}}\right), 
\end{align}
for all $\Gamma$, where $s\equiv q+p(1-\Gamma)$.

The vertical shear shear follows as
\begin{align}
  r\frac{\p\Omega^2}{\p z} = \frac{sz}{\Gamma r}  \frac{\OmK^2}{
    \left(1+z^2/r^2\right)^{3/2}} . \label{vertical_shear_ex} 
\end{align}
For $\Gamma = 1$, notice that $s/\Gamma = q$ is the only disk parameter that affects vertical shear.
Vertical shear increases linearly with height near the mid plane, and $\p_z \Omega^2$ is maximized at
the smaller of $r/\sqrt{2}$ or $H_s$.

@@

 Perturbations take the form of a radial plane wave with arbitrary vertical dependence, 
 e.g.\
 \begin{align}
  \delta\rho \rightarrow \delta\rho(z)\exp{\left(\ii k_x x - \ii\upsilon
      t\right)},    
\end{align}
where $\upsilon = \omega + \ii \sigma$ is a (generally) complex frequency, 
with $\omega$ and $\sigma$ being
the real frequency and growth rate, respectively, and $k_x$ is a real 
radial wavenumber.  We take $k_x>0$ without loss of generality, 
assume $k_xr_0 \gg 1$ and neglect curvature terms. 
Henceforth all unperturbed fluid variables, including 
gradients such as $\p \Omega/ \p z$, refer to  equilibrium values at $(r_0, z)$.

We further define the perturbation variables  $W \equiv \delta P /\rho$ 
and $Q \equiv c_s^2\delta\rho/\rho$.  With this procedure the linearized
system of equations are
\begin{align}
  \ii\upsilon \frac{Q}{c_s^2}  &=  \ii k_x \delta v_x + \frac{d\dd
    v_z}{dz} + \frac{\p\ln{\rho}}{\p z}\delta v_z + \zeta
  \frac{\p\ln{\rho}}{\p r} \delta v_x,\label{lin_mass}\\
  \ii\upsilon \delta v_x  &= \ii k_x W - 2\Omega\delta v_y -
  \zeta\frac{1}{\rho}\frac{\p P}{\p r}\frac{Q}{c_s^2}\label{lin_vx}\\
   \ii \upsilon\delta v_y &= r_0\frac{\p \Omega}{\p z}\delta v_z +
  \frac{\kappa^2}{2\Omega}\delta v_x, \label{lin_vy}\\
   \ii\upsilon\delta v_z &= \frac{dW}{dz} +
  \frac{\p\ln{\rho}}{\p z}\left(W-Q\right), \label{lin_vz}\\
  \ii \upsilon W &= c_s^2\frac{\p\ln{\rho}}{\p z}\delta v_z +
  c_s^2\frac{\gamma}{\Gamma} \left(\ii k_x\delta v_x + \frac{d\delta
      v_z}{dz}\right) \notag\\
  &+ \frac{1}{t_c}\left(W - \frac{Q}{\Gamma}\right) +
  \zeta\frac{1}{\rho}\frac{\p P}{\p r}\delta v_x.\label{lin_energy}
  % -&\ii\upsilon W + \frac{1}{t_c}\left(W-\frac{Q}{\Gamma}\right) = -\ii\upsilon\frac{\gamma}{\Gamma} Q +
  % c_s^2\frac{d\ln{\rho}}{dz}\left(\frac{\gamma}{\Gamma}-1\right)\delta v_z.\label{lin_energy}
\end{align}
This is a set of ordinary differential equations (ODEs) in
$z$.  %boundary conditions for well posed problem 
Solutions to these 
are presented in \S\ref{numerical} and \S\ref{application}.   The coefficient $ \zeta = 1$ is 
introduced simply to label terms containing radial gradients of the equilibrium state, 
which again are evaluated at $(r_0, z)$. For clarity, hereafter we drop the subcript $0$.    
We will consider the effects of ignoring these radial gradients below.

>>>>

We further define a modified sound speed
\begin{align}
	c_s \equiv \sqrt{\Gamma P /\rho } , 
% c_s^2\equiv % \frac{dP}{d\rho} =
%  c_{s0}^2(r)\left(\frac{\rho}{\rho_0}\right)^{\Gamma-1} =
%  \frac{\Gamma P}{\rho},  
\end{align}
which in general differs from the  isothermal, $c_\mathrm{iso} =
c_s/\sqrt{\Gamma}$, and adiabatic $c_\mathrm{ad} =
c_s\sqrt{\gamma/\Gamma}$, sound speeds. 
We also introduce the characteristic scale-height, $H = c_s(r,0)/\OmK$
and disk aspect-ratio
%We take $c_{s0}(r)$ to be another power-law:  
%\begin{align}
%  c_{s0}^2(r)=c_{s00}^2\left(\frac{r}{r_0}\right)^q, 
%\end{align}
%where $c_{s00}\equiv c_{s0}(r_0)$  and $q$ is the temperature power-law index, and
%define   
\begin{align}
   h(r) \equiv %\frac{c_{s0}(r)}{v_k(r)} \equiv
  \frac{H}{r} \propto r^{(1+q)/2}
\end{align}
We are interested in thin disks with $ h \ll 1$.

\subsubsection{Density structure}
The equilibrium density field follows from vertical hydrostatic equilibrium (Eq.\ \ref{vert_equilibrium}).
The solution for $\Gamma\neq1$ is
\begin{align}\label{eqm_dens}
  &\left(\frac{\rho}{\rho_0}\right)^{\Gamma-1} = 1 +
  \frac{\left(\Gamma-1\right)}{ h^2}\left(\frac{1}{\sqrt{1+z^2/r^2}}-1\right).
\end{align}
Note that for $\Gamma > 1 +  h^2$ there is a disk surface $H_s$
where $\rho(r,H_s)=0$; and $H_s \simeq \sqrt{2/(\Gamma-1)}H$ for $h\ll 1$.
%For $h \ll 1$, this surface is located at $H_s \simeq \sqrt{2/(\Gamma-1)}H$. 
% Note that for $\Gamma\neq1$ the disk has a surface $z=H(r)$ such that
% $\rho(r,H)=0$. If we parametrize
% \begin{align}
%   H(r)\equiv h r,
% \end{align}
% then the imposed sound-speed and disk thickness are related by 
% \begin{align}
%   h^2(r) = \left[1-\frac{ h^2(r)}{\Gamma-1}\right]^{-2}-1. 
% \end{align}

The vertically isothermal case, $\Gamma = 1$, can be calculated either 
as a special case or as the $\Gamma \rightarrow 1$ limit of Eq. \ref{eqm_dens},    
\begin{subequations}\begin{align}
\lim_{\Gamma\to1}  \frac{\rho(r,z)}{\rho_0(r)} &=
  \exp{\left[\frac{1}{ h^2}\left(\frac{1}{\sqrt{1+z^2/r^2}}-1\right)\right]},   \\
  & \simeq \exp {\left( - {z^2 \over 2 H^2}\right)} \quad\text{for } |z|\ll r \label{rhoisothin}. 
  \end{align}\end{subequations}
Eq.\ \ref{rhoisothin} is the approximate form of the density field in the thin-disk limit.  
We will primarily focus on 
vertically isothermal disks, the relevant case for passively irradiated
PPDs \citep{chiang97}.

 \subsubsection{Rotation and vertical shear profiles}\label{vshear_def}
The equilibrium rotation field, $\Omega(r,z)$ follows from the density field and centrifugal balance (Eq.\ \ref{hori_equilibrium}),
giving 
\begin{align}\label{eqm_rot}
  \frac{\Omega^2(r,z)}{\OmK^2(r)}=1 +
  \frac{p+q}{\Gamma} h^2(r) 
  +\frac{s}{\Gamma} \left(1-\frac{1}{\sqrt{1+z^2/r^2}}\right), 
\end{align}
for all $\Gamma$, where $s\equiv q+p(1-\Gamma)$.  We also refer to the
epicyclic frequency $\kappa$ defined via
\begin{align}\label{kap2_def}
\kappa^2 = \frac{1}{r^3}\frac{\p}{\p r}\left(r^4\Omega^2\right). 
\end{align}

The vertical shear rate follows from Eq. \ref{eqm_rot}, 
\begin{align}
  r\frac{\p\Omega^2}{\p z} = \frac{sz}{\Gamma r}  \frac{\OmK^2}{
    \left(1+z^2/r^2\right)^{3/2}} . \label{vertical_shear_ex} 
\end{align}
For $\Gamma = 1$, notice that $s/\Gamma = q$ is the only disk parameter that affects vertical shear.
Vertical shear increases linearly with height near the midplane, and $|\p_z \Omega^2|$ is maximized at
$\mathrm{min}(r/\sqrt{2},H_s)$, which is expected to limit the maximum VSI growth rate. 
%the smaller of $r/\sqrt{2}$ or $H_s$.

@@

 Perturbations take the form of a radial plane wave with arbitrary vertical dependence, 
 e.g.\
 \begin{align}
  \delta\rho \rightarrow \delta\rho(z)\exp{\left(\ii k_x x - \ii\upsilon
      t\right)},    
\end{align}
where $\upsilon = \omega + \ii \sigma$ is a (generally) complex frequency, 
with $\omega$ and $\sigma$ being
the real frequency and growth rate, respectively, and $k_x$ is a real 
radial wavenumber.  We take $k_x>0$ without loss of generality, 
assume $k_xr_0 \gg 1$ and neglect curvature terms. 
Henceforth all unperturbed fluid variables, including 
gradients such as $\p_z \Omega$, refer to  equilibrium values at $(r_0, z)$.

We further define the perturbation variables  $W \equiv \delta P /\rho$ 
and $Q \equiv c_s^2\delta\rho/\rho$.  With this procedure the linearized
system of equations are
\begin{subequations}\label{lin_all}\begin{align}
  \ii\upsilon \frac{Q}{c_s^2}  &=  \ii k_x \delta v_x + \frac{d\dd
    v_z}{dz} + \frac{\p\ln{\rho}}{\p z}\delta v_z + \zeta
  \frac{\p\ln{\rho}}{\p r} \delta v_x,\label{lin_mass}\\
  \ii\upsilon \delta v_x  &= \ii k_x W - 2\Omega\delta v_y -
  \zeta\frac{1}{\rho}\frac{\p P}{\p r}\frac{Q}{c_s^2}\label{lin_vx}\\
   \ii \upsilon\delta v_y &= r_0\frac{\p \Omega}{\p z}\delta v_z +
  \frac{\kappa^2}{2\Omega}\delta v_x, \label{lin_vy}\\
   \ii\upsilon\delta v_z &= \frac{dW}{dz} +
  \frac{\p\ln{\rho}}{\p z}\left(W-Q\right), \label{lin_vz}\\
  \ii \upsilon W &= c_s^2\frac{\p\ln{\rho}}{\p z}\delta v_z +
  c_s^2\frac{\gamma}{\Gamma} \left(\ii k_x\delta v_x + \frac{d\delta
      v_z}{dz}\right) \notag\\
  &+ \frac{1}{t_c}\left(W - \frac{Q}{\Gamma}\right) +
  \zeta\frac{1}{\rho}\frac{\p P}{\p r}\delta v_x.\label{lin_energy}
  % -&\ii\upsilon W + \frac{1}{t_c}\left(W-\frac{Q}{\Gamma}\right) = -\ii\upsilon\frac{\gamma}{\Gamma} Q +
  % c_s^2\frac{d\ln{\rho}}{dz}\left(\frac{\gamma}{\Gamma}-1\right)\delta v_z.\label{lin_energy}
\end{align}\end{subequations}
This is a set of ordinary differential equations (ODEs) in
$z$.  %boundary conditions for well posed problem 
Solutions to these 
are presented in \S\ref{numerical} and \S\ref{application}.   The coefficient $ \zeta = 1$ is 
introduced simply to label terms containing radial gradients of the equilibrium state, 
which again are evaluated at $(r_0, z)$. For clarity, hereafter we drop the subcript $0$.    
We will consider the effects of ignoring these radial gradients below.
